Supporting Information:\\Evidence for black-swan events in animal populations

### 2015-07-22 text/ms.tex
Text edits (and unwrap)

Black swans are statistically improbable events that nonetheless occur---often with profound consequences\cite{taleb2007, sornette2009}. While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency\cite{meehl2004, katz2005, ipcc2012}, it remains unclear the extent to which ecological populations buffer or suffer from such extremes. We derived a model for estimating the frequency of black-swan events and applied it to abundance time series of 609 animal populations. We find strong evidence for black-swan events, but they are rare, occurring in 3--5\% of populations: most frequently for birds (4--8\%), mammals (4--6\%), and insects (2--3\%). Black swan events were predominantly (87\%) downward, implying that unexpected population crashes occur far more frequently than unexpected population increases. Black swan events are not explained by any life-history covariates, but tend to be driven by external perturbations such as climate or severe winters. Data and simulations reveal that black swan-events are more likely to be detected in longer and less noisy time series. Our results show that accurate forecasts of extinction risk require accounting for the possibility of black swans and the predominance of crashes over increases, or predicted abundance will underestimate the probability of crashes by a factor of XX--XX. In addition, the presence of black swans highlights the importance of developing flexible and robust conservation and management strategies, particularly as the frequency and magnitude of climate extremes increase over the next century.

@@

Here, we assess the frequency and magnitude of black-swan dynamics across time series of 609 populations from a wide array of taxonomic groups---mostly birds, mammals, insects, and fishes. We then identify characteristics of time series or life-history traits associated with the detection of black-swan events and verify known causes. To accomplish this, we develop a framework for identifying heavy-tailed (black-swan) process noise in population dynamics, i.e.~whether the largest stochastic jumps in abundance from one time step to the next are more extreme than typically seen with a lognormal distribution. Our framework allows for a range of population dynamic models, can incorporate observation uncertainty and skewness in process noise, and can be readily applied to abundance time series.

Specifically, we fit population dynamics models in which the process noise is drawn from a Student-t distribution\emph{.} By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are thereof evidence of black-swan events (Fig.~1). For example, at \(\nu = 3\), the probability of drawing a value more than five standard deviations below the mean is \(0.02\), whereas the probability of drawing such a value from a normal distribution is tiny (\(2.9\cdot10^-7\)). As \(\nu\) approaches infinity, the distribution approaches the normal distribution (Fig.~1).

We found strong, but rare, evidence for black-swan dynamics: most frequently for birds (7\%) followed by mammals (5\%), and insects (3\%) (Fig.~2, Table~S2). Black-swan events were taxonomically widespread, occurring in 38\% of taxonomic orders. Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, we found stronger evidence for black swans in insect populations than is apparent in Fig.~2---four of eight orders with the highest median probability of heavy tails were insect orders---however, there was considerable uncertainty in these estimates (Fig.~3a).

The majority of our heavy-tailed estimates were robust to alternative population models, observation error, and choice of Bayesian priors. Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates with or without density dependence, or modelled the population dynamics as Ricker-logistic instead of Gompertz (Fig.~S3). Introducing moderate observation error (CV = 0.2) slightly decreased the estimated prevalence of black-swan events (Fig.~S3). The strength of the prior on ν had little influence on estimates of black-swan dynamics (Fig.~S4). Our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy tailed events---especially given the length of the time series in our analysis (Figs S5, S6).

Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise but not clearly related to population growth rate, density dependence, or maximum lifespan (Figs 3b, 4). Longer time-series length was the strongest covariate of observing black-swan dynamics. For instance, the expected probability density suggesting black-swan dynamics was about 1.6 times greater for a population with 60 time steps compared to one with 30 time steps. However, the absolute change in probability with increased time series length was small (0.20 vs.~0.13 in this example, Fig.~4).

We examined all time series with published explanations of why black-swan events occurred (Table~S2). The majority (87\%) were downward black swans and involved a combination of multiple factors. For example, a synchronization of environmental- and predation-mediated population cycles is thought to have caused a downward black-swan event for a water vole (\emph{Arvicola terrestris}) population\cite{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\emph{Ardea cinerea}) abundance in the United Kingdom\cite{stafford1971} (Fig.~4c). Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted\cite{stafford1971}. Downward black swans were sometimes followed by upward black swans. For example, during a period of population crowding and nest shortages, a population of European shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United Kingdom, declined suddenly following a red tide event in 1968\cite{potts1980}. This freed quality nest sites for first-time breeders, productivity rapidly increased, and the population experienced a rapid upswing in abundance\cite{potts1980}.

@@

We chose a weakly informative exponential prior on \(\nu\) with rate parameter of \(0.01\) truncated at values above two---a slightly less informative prior than suggested by Fernandez and Steel\cite{fernandez1998}. We truncated the distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\) but constrains the sampling sufficiently to avoid drifting towards infinity. In any case, for \(\nu > 20\) the t distribution is almost indistinguishable from the normal distribution (Fig.~1). Based on the shape of the t distribution, we chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the probability of heavy-tailed dynamics. When categorizing a population as heavy or normal tailed, we applied a 0.5 probability threshold. In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~1e, h), the posterior will approximately match the prior (prior median \(= 71\), mean \(= 102\)) and a metric of \(Pr(\nu < 10) > 0.5\)) is unlikely to flag the population as heavy tailed.

@@

\textbf{Skewed Student-t distribution} We fit Gompertz models with skewed Student-t distributed process noise based on the probability density for the skewed t distribution described in Fernandez and Steel\cite{fernandez1998}. Their definition equates to the following probability density as implemented in King\cite{king2012}:

\textbf{Alternative priors} To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics, we refit our models with weaker and stronger priors. Our base model used a prior on \(\nu\) of Truncated-Exponential(0.02, min.\ = 2). For a weaker prior we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we used Truncated-Exponential(0.02, min.\ = 2) (Extended Data Fig.~\ref{fig:priors}). Note that the base and weaker priors are relatively flat within the region of \(\nu < 20\), which is the region we are mostly concerned about when categorizing populations as heavy- or thin-tailed. TODO - note Gelman 1/x prior.

@@

\textbf{Alternative population models} We fit alternative population models to test if four key phenomena systematically changed our conclusions The range of percentages of black swans by taxonomic class cited in the abstract are based on lower and upper limits across our main Gompertz model and these four alternative models.

@@

We found that we could consistently and precisely recover median posterior estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge 200\)) (Extended Data Fig.~\ref{fig:sim-nu} upper panels). At smaller samples we could still usually qualitatively distinguish heavy from not-heavy tails, but the model tended to underestimate how heavy the tails were. At the same time, at smaller sample sizes, the model tended to overestimate how large the scale parameter was (Extended Data Fig.~\ref{fig:sim-nu} lower panels).

@@

\textbf{Covariates of heavy-tailed dynamics} We fit a multilevel beta regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)), to investigate potential covariates of heavy-tailed dynamics. The beta distribution is useful when response data range on a continuous scale between zero and one\cite{ferrari2004}. We used a logit link function as is typically used in logistic regression. The model was as follows:
\begin{align}
\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]}
  + \alpha^\mathrm{species}_{l[i]}
  + X_i \beta),
  \: \text{for } i = 1, \dots, 617\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,
\end{align}
where \(A\) and \(B\) represent the beta distribution shape parameters; \(\mu_i\) represents the predicted value for population \(i\), class \(j\), order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the dispersion parameter; and \(X_i\) represents a vector of predictors (such as lifespan) for population \(i\) with associated \(\beta\) coefficients. The intercepts are allowed to vary from the overall intercept \(\alpha\) by taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order (\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with standard deviations \(\sigma_{\alpha \;  \mathrm{class}}\), \(\sigma_{\alpha \; \mathrm{order}}\), and \(\sigma_{\alpha  \; \mathrm{species}}\). Where possible, we also allowed for error distributions around the predictors by incorporating the standard deviation of the posterior samples for the Gompertz parameters \(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal distributions (not shown in the above equation).

@@

We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0, 10)\) on the global intercept \(\alpha\), \(\mathrm{Half\mhyphen Cauchy}(0, 2.5)\) on all standard deviation parameters, \(\mathrm{Half\mhyphen Cauchy}(0, 10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and \(\mathrm{Cauchy}(0, 2.5)\) on all other parameters\cite{gelman2006c, gelman2008d}. Compared to normal priors, the Cauchy priors concentrate more probability density around expected parameter values while allowing for a higher probability density far into the tails, thereby allowing the data to dominate the posterior more strongly if it disagrees with the prior. Our conclusions were not qualitatively changed by using uniform priors. We fit our models with 5000 total iterations per chain, 2500 warm-up iterations, four chains, and discarding every second sample to save memory. We checked for chain convergence visually and with the same criteria as before (\(\widehat{R} < 1.05\) and \(n_\mathrm{eff} >200\) for all parameters).

To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig.~3a), we fit a separate multilevel model with the same structure but with only \(\log\) time-series length as a predictor. (In this case, we did not want to control for intrinsic population characteristics such as density dependence.) Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for \(\alpha\), \(\alpha^\mathrm{class}_j\), and \(\alpha^\mathrm{order}_k\).

@@

\caption{
  \textbf{Probability density of the Bayesian priors for the Gompertz models.} From
  left to right and then top to bottom: (1) per capita growth rate at
  $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale
  parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen
    Cauchy} (0, 2.5)$; (3) t-distribution degrees of freedom parameter: $\nu
  \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4)
  AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated
    \mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown
  is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
  The $\nu$ panel also shows two alternative priors: a weaker prior $\nu \sim
  \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a
  stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02,
  \mathrm{min.} = 2)$. The inset panel shows the same data but with a
  log-transformed x axis. Note that the base and weaker priors are relatively
  flat within the region of $\nu < 20$ that we are concerned with. }

@@

\caption{Testing the ability to estimate $\nu$ (top
  panels) and the scale parameter of the process deviations (bottom panels)
  for a given number of samples (columns) drawn from a distribution with a
  given true $\nu$ value (rows). The red lines indicate the true population
  value. When a small number of samples are drawn there may not be samples
  sufficiently far into the tails to recapture the true $\nu$ value; however,
  heavy tails are still distinguished from normal tails in most cases, even
  with only 25 or 50 samples.}

@@

\caption{Simulation testing the Gompertz estimation model when the process
deviation draws were chosen so that $\nu$ could be estimated close to the true
value outside the full population model (``effective $\nu$'' within a CV of 0.2
of specified $\nu$). Upper panels show the distribution of median
$\widehat{\nu}$ across 20 simulation runs. Lower panels show the distribution
of Pr($\nu < 10$) across 20 simulation runs. We ran the simulations across
three population (``true'') $\nu$ values (3, 5, and $1\cdot 10^9$, i.e.\
approximately normal) and four scenarios: (1) 100 time steps and no observation
error, (2) 50 time steps and no observation error, (3) 50 time steps and
observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4)
50 time steps with observation error in which the quantity of observation error
was assumed known. Within each scenario the dots represent stochastic draws
from the true population distributions combined with model fits. Underlayed
boxplots show the median, interquartile range, and $1.5$ times the
interquartile range. }

>>>>

Black swans are statistically improbable events that nonetheless occur---often with profound consequences\cite{taleb2007, sornette2009}. While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency\cite{meehl2004, katz2005, ipcc2012}, it remains unclear the extent to which ecological populations buffer or suffer from such extremes. We derived a model for estimating the frequency of black-swan events and applied it to abundance time series of 609 animal populations. We find strong evidence for black-swan events, but they are rare, occurring in 3--5\% of populations: most frequently for birds (4--8\%), mammals (4--6\%), and insects (2--3\%). Black swan events were predominantly (\percBSDown \%) downward, implying that unexpected population crashes occur far more frequently than unexpected population increases. Black swan events are not explained by any life-history covariates, but tend to be driven by external perturbations such as climate or severe winters. Data and simulations reveal that black swan-events are more likely to be detected in longer and less noisy time series. Our results show that accurate forecasts of extinction risk require accounting for the possibility of black swans and the predominance of crashes over increases, or predicted abundance will underestimate the probability of crashes by a factor of XX--XX. The presence of black swans highlights the importance of developing flexible and robust conservation and management strategies, particularly as the frequency and magnitude of climate extremes increase over the next century.

@@

Here, we assess the frequency and magnitude of black-swan dynamics across time series of 609 populations from a wide array of taxonomic groups---mostly birds, mammals, insects, and fishes (Extended Data Table~1). We then identify characteristics of time series or life-history traits associated with the detection of black-swan events and verify known causes. To accomplish this, we develop a framework for identifying heavy-tailed (black-swan) process noise in population dynamics, i.e.~whether the largest stochastic jumps in abundance from one time step to the next are more extreme than typically seen with a lognormal distribution. Our framework allows for a range of population dynamic models, can incorporate observation uncertainty and skewness in process noise, and can be readily applied to abundance time series.

Specifically, we fit population dynamics models in which the process noise is drawn from a Student-t distribution. By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are thereof evidence of black-swan events (Fig.~1). For example, at \(\nu = 3\), the probability of drawing a value more than five standard deviations below the mean is \(0.02\), whereas the probability of drawing such a value from a normal distribution is tiny (\(2.9\cdot10^-7\)). As \(\nu\) approaches infinity, the distribution approaches the normal distribution (Fig.~1). TODO CITE Extended Data Fig.~\ref{fig:priors}

We found strong, but rare, evidence for black-swan dynamics: most frequently for birds (7\%) followed by mammals (5\%), and insects (3\%) (Fig.~2). Black-swan events were taxonomically widespread, occurring in 38\% of taxonomic orders. Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, we found stronger evidence for black swans in insect populations than is apparent in Fig.~2---four of eight orders with the highest median probability of heavy tails were insect orders---however, there was considerable uncertainty in these estimates (Fig.~3a).

The majority of our heavy-tailed estimates were robust to alternative population models, observation error, and choice of Bayesian priors. Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates with or without density dependence, or modelled the population dynamics as Ricker-logistic instead of Gompertz (Extended Data Fig.~\ref{fig:alt}). Introducing moderate observation error (CV = 0.2) slightly decreased the estimated prevalence of black-swan events (Extended Data Fig.~\ref{fig:alt}). The strength of the prior on \(\nu\) had little influence on estimates of black-swan dynamics (Extended Data Fig.~\ref{fig:alt-priors}). Our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy tailed events---especially given the length of the time series in our analysis (Extended Data Figs \ref{fig:sim-nu}, \ref{fig:sim-prob}).

Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise but not clearly related to population growth rate, density dependence, or maximum lifespan (Fig.~3b, Extended Data Fig.~\ref{fig:correlates}). Longer time-series length was the strongest covariate of observing black-swan dynamics. For instance, the expected probability density suggesting black-swan dynamics was about 1.6 times greater for a population with 60 time steps compared to one with 30 time steps. However, the absolute change in probability with increased time series length was small (0.20 vs.~0.13 in this example, Extended Data Fig.~\ref{fig:correlates}).

We examined all time series with published explanations of why black-swan events occurred (Extended Data Table~2). The majority (\percBSDown \%) were downward black swans and involved a combination of multiple factors. For example, a synchronization of environmental- and predation-mediated population cycles is thought to have caused a downward black-swan event for a water vole (\emph{Arvicola terrestris}) population\cite{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\emph{Ardea cinerea}) abundance in the United Kingdom\cite{stafford1971} (Fig.~4c). Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted\cite{stafford1971}. Downward black swans were sometimes followed by upward black swans. For example, during a period of population crowding and nest shortages, a population of European shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United Kingdom, declined suddenly following a red tide event in 1968\cite{potts1980}. This freed quality nest sites for first-time breeders, productivity rapidly increased, and the population experienced a rapid upswing in abundance\cite{potts1980}.

@@

We chose a weakly informative exponential prior on \(\nu\) with rate parameter of \(0.01\) truncated at values above two---a slightly less informative prior than suggested by Fernandez and Steel\cite{fernandez1998}. We truncated the distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\) but constrains the sampling sufficiently to avoid drifting towards infinity. In any case, for \(\nu > 20\) the t distribution is almost indistinguishable from the normal distribution (Fig.~1). Based on the shape of the t distribution, we chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the probability of heavy-tailed dynamics. When categorizing a population as heavy or normal tailed, we applied a 0.5 probability threshold. In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~1e, h), the posterior will approximately match the prior (prior median \(= 71\), mean \(= 102\)) and a metric of Pr\((\nu < 10) > 0.5\)) is unlikely to flag the population as heavy tailed.

@@

\textbf{Alternative priors} To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics, we refit our models with weaker and stronger priors. Our base model used a prior on \(\nu\) of Truncated-Exponential(0.02, min.\ = 2). For a weaker prior we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we used Truncated-Exponential(0.02, min.\ = 2) (Extended Data Fig.~\ref{fig:priors}). Note that the base and weaker priors are relatively flat within the region of \(\nu < 20\), which is the region we are mostly concerned about when categorizing populations as heavy- or thin-tailed. We furthermore considered a Uniform(\(0, 1\)) prior on \(1/\nu\) as used in Gelman \emph{et al.}\cite{gelman2014}, but this prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\) than the truncated exponential priors considered here (not shown).

@@

\textbf{Alternative population models} We fit alternative population models to test if four key phenomena systematically changed our conclusions. The range of percentages of black swans by taxonomic class cited in the abstract are based on lower and upper limits across our main Gompertz model and these four alternative models.

@@

We found that we could consistently and precisely recover median posterior estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge 200\)) (Extended Data Fig.~\ref{fig:sim-nu} upper panels). At smaller samples we could usually qualitatively distinguish heavy from not-heavy tails, but the model tended to underestimate how heavy the tails were. At the same time, at smaller sample sizes, the model tended to overestimate how large the scale parameter was (Extended Data Fig.~\ref{fig:sim-nu} lower panels).

@@

\textbf{Covariates of heavy-tailed dynamics} We fit a multilevel beta regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)), to investigate potential covariates of heavy-tailed dynamics. We obtained maximum lifespan and body-size data from Brook \emph{et al.}\cite{brook2006a}. The beta distribution is useful when response data range on a continuous scale between zero and one\cite{ferrari2004}. We used a logit link function as is typically used in logistic regression. The model was as follows:
\begin{align}
\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]}
  + \alpha^\mathrm{species}_{l[i]}
  + X_i \beta),
  \: \text{for } i = 1, \dots, 617\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,
\end{align}
where \(A\) and \(B\) represent the beta distribution shape parameters; \(\mu_i\) represents the predicted value for population \(i\), class \(j\), order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the dispersion parameter; and \(X_i\) represents a vector of predictors (such as lifespan) for population \(i\) with associated \(\beta\) coefficients. The intercepts are allowed to vary from the overall intercept \(\alpha\) by taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order (\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \; \mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\). Where possible, we also allowed for error distributions around the predictors by incorporating the standard deviation of the posterior samples for the Gompertz parameters \(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal distributions (not shown in the above equation).

@@

We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0, 10)\) on the global intercept \(\alpha\), \(\mathrm{Half\mhyphen Cauchy}(0, 2.5)\) on all standard deviation parameters, \(\mathrm{Half\mhyphen Cauchy}(0, 10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and \(\mathrm{Cauchy}(0, 2.5)\) on all other parameters\cite{gelman2006c, gelman2008d}. Compared to normal priors, the Cauchy priors concentrate more probability density around expected parameter values while allowing for a higher probability density far into the tails, thereby allowing the data to dominate the posterior more strongly if it disagrees with the prior. We fit our models with 5000 total iterations per chain, 2500 warm-up iterations, four chains, and discarding every second sample. We checked for chain convergence visually and with the same criteria as before (\(\widehat{R} < 1.05\) and \(n_\mathrm{eff} >200\) for all parameters).

To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig.~3a), we fit a separate multilevel model with the same structure but with only \(\log\) time-series length as a predictor---in this case, we did not want to control for intrinsic population characteristics such as density dependence. Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for \(\alpha\), \(\alpha^\mathrm{class}_j\), and \(\alpha^\mathrm{order}_k\).

\textbf{Skewed Student-t forecasts} To formally evaluate the apparent skewness of heavy-tailed process noise we fit Gompertz models with skewed Student-t distributed process noise based on the probability density for the skewed t distribution described in Fernandez and Steel\cite{fernandez1998}. Their definition equates to the following probability density as implemented in King\cite{king2012}:

TODO \ldots{} we fit these\ldots{} and compare 5-year forecasts to standard Gompertz with normal error. We project all with median nu \textless{} X and calculate the ratio of the lower 99\% credible interval between the two models.. increased number of samples because getting far into tails, aggregate posterior.

@@

\caption{\textbf{Probability density of the Bayesian priors for the Gompertz models.} From left to right and then top to bottom: (1) per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$. The $\nu$ panel also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$. The inset panel shows the same data but with a log-transformed x axis. Note that the base and weaker priors are relatively flat within the region of $\nu < 20$ that we are concerned with. }

@@

\caption{\textbf{Testing the ability to estimate $\nu$ (top panels) and the scale parameter of the process deviations (bottom panels) for a given number of samples (columns) drawn from a distribution with a given true $\nu$ value (rows).} The red lines indicate the true population value. When a small number of samples are drawn there may not be samples sufficiently far into the tails to recapture the true $\nu$ value; however, heavy tails are still distinguished from normal tails in most cases, even with only 25 or 50 samples.}

@@

\caption{\textbf{Simulation testing the Gompertz estimation model when the process deviation draws were chosen so that $\nu$ could be estimated close to the true value outside the full population model} (``effective $\nu$'' within a CV of 0.2 of specified $\nu$). Upper panels show the distribution of median $\widehat{\nu}$ across 20 simulation runs. Lower panels show the distribution of Pr($\nu < 10$) across 20 simulation runs. We ran the simulations across three population (``true'') $\nu$ values (3, 5, and $1\cdot 10^9$, i.e.\ approximately normal) and four scenarios: (1) 100 time steps and no observation error, (2) 50 time steps and no observation error, (3) 50 time steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4) 50 time steps with observation error in which the quantity of observation error was assumed known. Within each scenario the dots represent stochastic draws from the true population distributions combined with model fits. Underlayed boxplots show the median, interquartile range, and $1.5$ times the interquartile range.}
